Efficient and Scalable Bipartite Matching with Fast Beta Linkage (fabl)

### 2023-07-17 ba_resubmit/kundinger_resubmit2.tex
attempting to clarify some of brian's questions.

The independent pairwise matching assumption from \cite{fellegi_theory_1969} is popular mainly for its mathematical simplicity. However, in many situations, there are no duplications within a data file, meaning that each record in one file should be linked with at most one other record in the other file. Thus, when the procedure results in many-to-one matches, some of these links must be false. To address this potential problem, data analysts typically use an additional post-processing step, for example, applying an optimization routine that turns the list of linked records into a bipartite matching \citep{jaro1989} \textcolor{teal}{or using decision theoretic procedures, such as those of \cite{sadinle_bayesian_2017} and \cite{steorts_bayesian_2016}. For example, the decision theoretic work of \cite{steorts_bayesian_2016}, ensures that the resulting records are bipartite (or rather transitive).}

@@

In this article, we propose fast beta linkage (\texttt{fabl}), which extends the \texttt{BRL} model for increased efficiency and scalability. {\color{red}{We use independent priors for the matching status of each record, inducing a dependency that is stronger than the record pair independence in the original \cite{fellegi_theory_1969} but weaker that the one-to-one requirement of \cite{sadinle_bayesian_2017}.}} \textcolor{teal}{This independence assumption allows for a scalable update of the Gibbs sampler, contrasting sequential updates of \cite{sadinle_bayesian_2017}. We employ the technique of \cite{steorts_bayesian_2016} in order to ensure our method after record linkage is bipartite.}
This \textcolor{teal}{discovery} allows us to (1) employ hashing techniques that speed up calculations and reduce computational costs, (2) compute the pairwise record comparisons over large data files via parallel computing, and (3) reduce memory costs through what we call storage efficient indexing. These contributions allow \texttt{fabl} to perform record linkage on much larger data files than previous Bayesian Fellegi-Sunter models at significantly increased speed with similar levels of accuracy. In particular, computation time under \texttt{BRL} grows quadratically, with the size of each data file, while computation time under \texttt{fabl} grows linearly, only with the size of the smaller data file.

@@

\cite{sadinle_bayesian_2017} presents a Bayesian version of the model in \eqref{eqn:fs_model} through \eqref{eqn:lambda_dist}.  He uses uniform Dirichlet prior distributions for each $\bm{m}_f$ and $\bm{u}_f$.  As the prior distriburion for  $\bm{Z}$, he proposes the ``beta distribution for bipartite matching."  This assigns a Bernoulli distribution for the indicator that a record in $B$ has a match in $A$, that is, $I(Z_j \leq n_A) \sim \text{Bernoulli}(\pi)$.  Additionally, $\pi \sim \text{Beta}(\alpha_{\pi}, \beta_{\pi})$. It follows that the number of records in $B$ that have matches, denoted $n_{AB}(\bm{Z}) = \sum_{j=1}^{n_B} I(Z_j \leq n_A)$, is distributed according to a $\text{Beta-Binomial}(n_B, \alpha_{\pi}, \beta_{\pi})$. Conditioning on the set of records in $B$ that have matches, formally denoted $\{I(Z_j \leq n_A)\}_{j=1}^{n_B}$, all $n_A ! / (n_A - n_{AB}(\bm{Z}))!$ bipartite matchings are taken to be equally likely. Thus, the prior distribution used by \cite{sadinle_bayesian_2017} is given by
\begin{align}
\label{eqn:sadinle_prior}
p(\bm{Z}|\alpha_{\pi}, \beta_{\pi}) = \frac{(n_A - n_{AB}(\bm{Z}))!}{n_A !}\frac{\text{B}(n_{AB}(\bm{Z}) + \alpha_{\pi}, n_B - n_{AB}(\bm{Z}) + \beta_{\pi})}{\text{B}(\alpha_{\pi}, \beta_{\pi})},
\end{align}
where $\text{B}(\cdot, \cdot)$ represents the Beta function. This prior strictly enforces one-to-one matching, inducing a Gibbs sampler that removes previously matched records from the set of candidate records when sampling each $Z_j$. This makes the sampler inherently serial, which can be slow when working on linkage tasks with more than a few thousand records.

%\section{Review of Prior Work}
%\label{sec:review-of_prior-work}
%
%%Consider two data files $A$ and $B$ containing records $\{x_{1i}\}_{i=1}^{n_A}$ and $\{x_{2j}\}_{j=1}^{n_B}$ respectively.
%
%Consider two data files $A$ and $B$, consisting of records $\{A_i: i = 1, \ldots, n_A\}$ and $\{B_j: j = 1, \ldots, n_B\}$. Suppose the files have $F$ fields in common to be used for linkage, and that these fields take on values $A_{if}$ and $B_{jf}$ respectively. Without loss of generality, denote files such that $n_A \geq n_B$. Under bipartite matching, the set of matches across data files can be represented in two equivalent ways. First, we may use a matrix $\Delta \in \{0, 1\}^{n_A \times n_B}$, where
%\begin{align}
%	\Delta_{ij} =
%	\begin{cases}
%		1, & \text{if records}\;  A_i \; \text{and}\; B_j \; \text{refer to the same entity}; \\
%		0, & \text{otherwise}.\\
%	\end{cases}
%\end{align}
%This sparse matrix representation can become cumbersome for large linkage tasks. More compactly, bipartite matching also can be viewed as a labeling $\bm{Z} = (Z_1, \ldots, Z_{n_B})$ for the records in $B$ such that 
%\begin{align}
%	Z_{j} =
%	\begin{cases}
%		i, & \text{if records}\;  A_i \; \text{and}\; B_j  \; \text{refer to the same entity}; \\
%		n_A + j, & \text{if record}\;  B_j \; \text{does not have a match in}\; A. \\
%	\end{cases}
%\end{align}
%We can go back and forth between the two using $\Delta_{ij} = I(Z_j = i),$ where $I(\cdot) = 1$ when the expression inside the parentheses is true, and $I(\cdot) = 0$ otherwise. 
%

%Denote the set of matches by $\bm{M} = \{(i,j): i \leq n_A, j \leq n_B, \Delta_{ij} = 1\}$, and the set of nonmatches by 
%$\bm{U} =  \{(i,j): i \leq n_A, j \leq n_B, \Delta_{ij} = 0\}.$ The record linkage task can be viewed as identifying  $\bm{M}$ and  $\bm{U}.$ We refer to record pairs that are estimated as matches as ``links'' and to record pairs that are estimated as nonmatches as ``non-links.''
%
%%\subsection{Comparison Vectors}
%%\label{comparison-vectors}
%

%
%Intuitively, matching records (those that refer to the same entity) should be similar; records that are nonmatching should be dissimilar. \cite{fellegi_theory_1969} proposed encoding this using a comparison vector $\gamma_{ij}$ computed for each record pair $(i,j)$ in $A \times B.$ Denote the number of criteria for comparing records by $F$, such that $\gamma_{ij} = (\gamma_{ij}^1, \ldots, \gamma_{ij}^F).$ {\color{red}{We use $\Gamma_{ij}$ when referring to the vector as a random variable, and $\gamma_{ij}$ when referring to observed data. We collect the vectors $\gamma_{ij}$ as the comparison matrix $\gamma \in \mathbb{R}^{n_A n_B \times F}$}}.
%
%%Frequently, $\gamma_{ij}$ consists of one comparison for each feature shared between the two data files. 
%
%The simplest way to compare any particular feature for two records is to check for exact agreement, and this is commonly used for categorical features. For example, if zip code is linking feature $f$, we can set $\gamma_{ij}^f=1$ when the zip codes for records $A_i$ and $B_j$ agree exactly, and set $\gamma_{ij}^f=2$ when they do not. For numerical features, we can use absolute difference between the two feature values. For example, if age is linking field $f$, we can set $\gamma_{ij}^f = 1$ when the ages for records $A_i$ and $B_j$ match exactly, $\gamma_{ij}^f = 2$ when the ages for records $A_i$ and $B_j$ are within one year but not equal, and $\gamma_{ij}^f = 3$ when the ages are two or more years apart. For text features, such as names, we can use string distance metrics such as Levenstein or Jaro-Winkler distance \citep{cohen2003comparison}. We then set thresholds that allow us to represent comparisons through discrete levels of disagreement \citep{bilenko2006riddle, elmagarmid_duplicate_2007}. 
%
%More generally, let $\mathcal{S}_f(i,j)$ denote a similarity measure for feature $f$ of records $A_i$ and $B_j,$ where the range of $\mathcal{S}_f$ can be divided into $L_f$ intervals denoted by $I_{f1}, \ldots, I_{fL_f}$. Here, $I_{f1}$ represents the highest level of agreement (including complete agreement) and $I_{fL_f}$ represents the highest level of disagreement (including complete disagreement). Thus, we can construct comparison vectors such that $\gamma_{ij}^f = l \; \text{if} \; \mathcal{S}_f(i,j) \in I_{fl}.$ The choices of $I_{fl}$ are application specific, as we discuss in the simulation and case studies.
%
%{\color{red}{
%In the construction of comparison vectors, it is common to encounter missing information in record $A_i$ or $B_j$. As a result, the comparison vector $\mathbf{\gamma}_{ij}$ will have missing values. We assume that this missingness occurs completely at random (MCAR, per \cite{LittleRubin2002}). To notate a missing value in any $\gamma_{ij}^f$, we use $I_{obs}(\gamma_{ij}^f)=1$ when $\gamma_{ij}^f$ is observed and $I_{obs}(\gamma_{ij}^f)=0$ otherwise. With the MCAR assumption, we can marginalize over the missing data, and do all computation simply using the observed data. }
%}
%
%%For more complex measurements, we can take into account partial agreement to more richly characterize the comparison;  
%
%%
%%In this paper, we use hashing and a new technique called storage efficient indexing to increase the scale of the linkage tasks we can undertake without blocking. This is useful when there is no reliable blocking field available, or one desires estimates of model parameters for the entire sample in question. In practice, \texttt{fabl} can be combined with blocking, but all derivations, simulations, and case studies are presented here without blocking. 
%
%\subsection{Fellegi-Sunter Models}
%\label{fellegi-sunter}
%
%The seminal \cite{fellegi_theory_1969} model employs two independence assumptions: first, that comparison vectors are conditionally independent given their matching status, and second, that the matching status of the record pairs are independent. Using these assumptions, \cite{winkler_state_1999}, \cite{jaro1989}, \cite{larsen_2001}, \cite{enamorado2019using} and others model the comparison data through mixture models of the form
%\begin{subequations}
%\begin{align}
%	&\Gamma_{ij} = \gamma_{ij} \mid \Delta_{ij} = 1 \stackrel{iid}{\sim} \mathcal{M}(\bm{m}), \label{eqn:fs_model} \\
%	&\Gamma_{ij} = \gamma_{ij} \mid \Delta_{ij} = 0  \stackrel{iid}{\sim} \mathcal{U}(\bm{u}), \label{eqn:m_dist}\\
%	&\Delta_{ij}   \stackrel{iid}{\sim} \text{Bernoulli}(\lambda). \label{eqn:u_dist}
%\end{align}
%\end{subequations}
%Here, $\mathcal{M}$ and $\mathcal{U}$ are the distributions for matching and nonmatching record pairs, $\bm{m}$ and $\bm{u}$ are their respective sets of parameters, and $\lambda$ is the marginal probability that a record pair is a match. When using comparison vectors with discrete agreement levels, $\mathcal{M}$ and $\mathcal{U}$ are collections of independent multinomial distributions for each linkage feature. Accordingly, $\bm{m} = (\bm{m}_1, \ldots, \bm{m}_F)$, where $\bm{m}_f = (m_{f1}, \ldots, m_{fL_f})$ and $m_{fl} = p(\gamma_{ij}^f = l|\Delta_{ij} = 1)$ for all fields $f$ and agreement levels $l$. The $\bm{u}$ parameters are defined similarly, with $u_{fl} = p(\gamma_{ij}^f = l|\Delta_{ij} = 0)$.
%
%
%%This prompted \cite{jaro1989} to formulate an optimization problem that produces a bipartite matching maximizing the sum of the Fellegi-Sunter weights among matched pairs.
%
%%These unknown parameters are often estimated using the EM algorithm \cite{winkler_state_1999}. 
%
%%The Fellegi-Sunter method then uses thresholds $T_m$ and $T_u$ such that all record pairs with $w_{ij} > T_m$ are declared as links, and all those with $w_{ij} < T_u$ are declared as non-links (with those such that $T_u < w_{ij} < T_m$ left undetermined and subject to clerical review). This is a problem in practice as transitive closures can be violated. Since each $w_{ij}$ is calculated independently, there can be situations in which both $w_{ij} > T_m$ and $w_{i'j} > T_m$, so both $(i,j)$ and $(i', j)$ are declared as links, even though such a matching is not allowed by assumption. 
%
%%To address this issue, \cite{jaro1989} proposed solving the following linear sum assignment problem:
%%\begin{align}
%%	\label{eqn:jaro}
%%	&\max_{\Delta} \sum_{i=1}^{n_A} \sum_{j=1}^{n_B} w_{ij} \Delta_{ij} 
%%	\quad \text{subject to} \quad \Delta_{ij} \in \{0,1\}; \\
%%	& \quad \quad \sum_{i=1}^{n_A}  \Delta_{ij}  \leq 1, j=1,2, \ldots n_B; \; \text{and} \notag\\
%%	& \quad \quad\sum_{i=1}^{n_B}  \Delta_{ij}  \leq 1, i=1,2, \ldots n_A. \notag
%%\end{align}
%%
%%The solution to this problem, for which several simple algorithms exist, is a bipartite matching that maximizes the sum of the Fellegi-Sunter weights among matched pairs. Jaro provided no theoretical justification for using this approach; however \cite{sadinle_bayesian_2017} recently showed that under certain conditions, (\ref{eqn:jaro}) is the maximum likelihood estimate for bipartite matching. 
%
%%The comparison data $\gamma_{ij}$ are not sufficient to determine whether a record is a match or nonmatch because of errors that naturally occur in the data. This motivated \cite{fellegi_theory_1969} to consider the likelihood ratio
%%\begin{align}
%%	\label{eqn:wts}
%%	w_{ij} = \frac{p(\gamma_{ij} \mid \Delta_{ij} = 1)}{p(\gamma_{ij} \mid \Delta_{ij} = 0)}
%%\end{align}
%%as a weight to estimate if record pairs are a match or nonmatch. This ratio will be large when there is strong evidence of the pair being a match, and small otherwise. 
%%
%%In practice, $p(\cdot \mid \Delta_{ij} = 1)$ and $p(\cdot \mid \Delta_{ij} = 0)$ are unknown to the modeler, and thus must be estimated. 
%%
%%%is the set of parameters $u_{fl} = p(\gamma_{ij}^f = l|Z_j \neq i)$ for the nonmatching pairs.
%%
%%After estimating $\bm{m}$ and $\bm{u}$, the Fellegi-Sunter method uses thresholds $T_m$ and $T_u$ such that all record pairs with $w_{ij} > T_m$ are declared as links, and all those with $w_{ij} < T_u$ are declared as non-links (with those such that $T_u < w_{ij} < T_m$ left undetermined and subject to clerical review). This is a problem in practice as transitive closures can be violated. Since each $w_{ij}$ is calculated independently, there can be situations in which both $w_{ij} > T_m$ and $w_{i'j} > T_m$, so both $(i,j)$ and $(i, j')$ are declared as links, even though such a matching is not allowed by assumption. 
%
%%\subsection{Beta Record Linkage Model}
%%\label{BRL}
%
%%To estimate a one-to-one matching without using a post-processing step, \cite{sadinle_bayesian_2017} incorporated this constraint directly into the model through a prior distribution over the matching space. 
%
%To estimate a one-to-one matching without using a post-processing step, \cite{sadinle_bayesian_2017} incorporates this constraint directly into the model through a Bayesian framework. In addition to {\color{red}{uniform}} Dirichlet priors for the $\bm{m}_f$ and $\bm{u}_f$ parameters, he proposes the ``beta distribution for bipartite matching" for the linkage parameter $\bm{Z}$. He assigns a prior distribution for the probability of the indicator that a record in $B$ has a match in $A$, so that $I(Z_j \leq n_A) \sim \text{Bernoulli}(\pi)$, where $\pi$ itself is taken to be distributed $\text{Beta}(\alpha_{\pi}, \beta_{\pi})$. It follows that the number of records in $B$ that have matches, denoted $n_{AB}(\bm{Z}) = \sum_{j=1}^{n_B} I(Z_j \leq n_A)$, is distributed according to a $\text{Beta-Binomial}(n_B, \alpha_{\pi}, \beta_{\pi})$. Conditioning on the set of records in $B$ that have matches, formally denoted $\{I(Z_j \leq n_A)\}_{j=1}^{n_B}$, all $n_A ! / (n_A - n_{AB}(\bm{Z}))!$ bipartite matchings are taken to be equally likely. Thus, the prior is given by
%\begin{align}
%\label{eqn:sadinle_prior}
%p(\bm{Z}|\alpha_{\pi}, \beta_{\pi}) = \frac{(n_A - n_{AB}(\bm{Z}))!}{n_A !}\frac{\text{B}(n_{AB}(\bm{Z}) + \alpha_{\pi}, n_B - n_{AB}(\bm{Z}) + \beta_{\pi})}{\text{B}(\alpha_{\pi}, \beta_{\pi})},
%\end{align}
%where $\text{B}(\cdot, \cdot)$ represents the Beta function. This prior strictly enforces one-to-one matching, inducing a Gibbs sampler that removes previously matched records from the set of candidate records when sampling each $Z_j$. This makes the sampler inherently serial, which can be slow when working on linkage tasks with more than a few thousand records.

>>>>

The independent pairwise matching assumption from \cite{fellegi_theory_1969} is popular mainly for its mathematical simplicity. However, in many situations, there are no duplications within a data file, meaning that each record in one file should be linked with at most one other record in the other file. Thus, when the procedure results in many-to-one matches, some of these links must be false. To address this potential problem, data analysts typically use an additional post-processing step, for example, applying an optimization routine that turns the list of linked records into a bipartite matching \citep{jaro1989} \textcolor{teal}{or using decision theoretic procedures, such as those of \cite{sadinle_bayesian_2017} and \cite{steorts_bayesian_2016}, which both ensure  bipartite (or rather transitive) matches.}

@@

In this article, we propose fast beta linkage (\texttt{fabl}), which extends the \texttt{BRL} model for increased efficiency and scalability. {\color{red}{We use independent priors for the matching status of each record, inducing a dependency that is stronger than the record pair independence in the original \cite{fellegi_theory_1969} but weaker that the one-to-one requirement of \cite{sadinle_bayesian_2017}.}} \textcolor{teal}{This independence assumption allows for a scalable update of the Gibbs sampler, contrasting sequential updates of \cite{sadinle_bayesian_2017}. We employ the decision theoretic technique of \cite{sadinle_bayesian_2017} and \cite{steorts_bayesian_2016} in order to ensure our method after record linkage is bipartite.}
This \textcolor{teal}{discovery} allows us to (1) employ hashing techniques that speed up calculations and reduce computational costs, (2) compute the pairwise record comparisons over large data files via parallel computing, and (3) reduce memory costs through what we call storage efficient indexing. These contributions allow \texttt{fabl} to perform record linkage on much larger data files than previous Bayesian Fellegi-Sunter models at significantly increased speed with similar levels of accuracy. In particular, computation time under \texttt{BRL} grows quadratically, with the size of each data file, while computation time under \texttt{fabl} grows linearly, only with the size of the smaller data file.

@@

\cite{sadinle_bayesian_2017} presents a Bayesian version of the model in \eqref{eqn:fs_model} through \eqref{eqn:lambda_dist}.  He uses uniform Dirichlet prior distributions for each $\bm{m}_f$ and $\bm{u}_f$.  As the prior distriburion for  $\bm{Z}$, he proposes the ``beta distribution for bipartite matching."  This assigns a Bernoulli distribution for the indicator that a record in $B$ has a match in $A$, that is, $I(Z_j \leq n_A) \sim \text{Bernoulli}(\pi)$.  Additionally, $\pi \sim \text{Beta}(\alpha_{\pi}, \beta_{\pi})$. It follows that the number of records in $B$ that have matches, denoted $n_{AB}(\bm{Z}) = \sum_{j=1}^{n_B} I(Z_j \leq n_A)$, is distributed according to a $\text{Beta-Binomial}(n_B, \alpha_{\pi}, \beta_{\pi})$. Conditioning on the set of records in $B$ that have matches, formally denoted $\{I(Z_j \leq n_A)\}_{j=1}^{n_B}$, all $n_A ! / (n_A - n_{AB}(\bm{Z}))!$ bipartite matchings are taken to be equally likely. Thus, the prior distribution used by \cite{sadinle_bayesian_2017} is given by
\begin{align}
\label{eqn:sadinle_prior}
p(\bm{Z}|\alpha_{\pi}, \beta_{\pi}) = \frac{(n_A - n_{AB}(\bm{Z}))!}{n_A !}\frac{\text{B}(n_{AB}(\bm{Z}) + \alpha_{\pi}, n_B - n_{AB}(\bm{Z}) + \beta_{\pi})}{\text{B}(\alpha_{\pi}, \beta_{\pi})},
\end{align}
where $\text{B}(\cdot, \cdot)$ represents the Beta function. This prior strictly enforces one-to-one matching, inducing a Gibbs sampler that removes previously matched records from the set of candidate records when sampling each $Z_j$. This makes the sampler inherently \textcolor{teal}{sequential}, which can be slow when working on linkage tasks with more than a few thousand records.
